---
title: "Lindsey's Climate Farmers Challenge"
output: html_notebook
---

The work involves the use and analysis of gridded data (usually remote sensing or
model products), including land cover, climate and soil. You will need to download
and visualize relationships between these over a chosen region.

Load required packages
```{r}
# List of packages to load
packages_today <- c("tidyverse","ecmwfr","ggspatial","rgdal")

# Load the packages
lapply(packages_today, library, character.only = TRUE)

setwd("C:/Users/linds/Dropbox/Documents (Selective Sync Conflict)/Job Applications/Climate Farmers/CF-Challenge")

```



_**Step 1: Region Selection**_

*Decide on a specific region to work with. This region should be large enough to
include various land use types but small enough to keep computation times
reasonable. A country or region within a country for example.
The region can be defined using simple longitude and latitude limits or using a
geospatial file (e.g. shape file or geojson). Use this region to filter downloaded data
in the next steps.*


First, I load the required packages for this step (sf and rnaturalearth) to access shape files of country and regional boundaries. I will work with the state of California, USA.
```{r}
# Load the packages

lapply(c("sf","rnaturalearth"), library, character.only = TRUE)

# Get the natural earth data for U.S. states
california_sf <- ne_states(country = "united states of america", returnclass = "sf") %>% 
  filter(us_states$name == "California") %>% st_as_sf()
```


I confirm the correct geometry of the shape file and write to the computer, encoding with "UTF-8", as in the original file. 
```{r}
#Use base R to view state outline
plot(california_sf$geometry)

#Save shapefile
st_write(california_sf, "california_sf.shp", driver = "ESRI Shapefile", encoding ="UTF-8")
```


I read back and plot the saved shape file to confirm correct writing to the drive.
```{r}
#Check file saved correctly by reading in shape file and plotting geometry
test_shp<-st_read("california_sf.shp")
plot(test_shp$geometry)
```


Finally, I plot on coordinates with ggplot2 to confirm the correct geographic position and begin building a graph for later visualisation.
```{r}
#Generate graph on coordinates to confirm position
ggplot(data = california_sf) +
  geom_sf() +
  ggtitle("California, USA") +
  theme(plot.title = element_text(hjust = 0.5)) +
  #xlab("") + ylab("") +
  geom_sf(color = "black", fill = "grey") 
```

_**Step 2: Data Acquisition**_

*Download data for your region for:
● Monthly evapotranspiration, temperature, and precipitation covering the
years 2000 to 2022 from (ERA5-Land):
https://cds.climate.copernicus.eu/#!/home
● Land cover classification maps for the year 2020 from Land Cover Classes
(Gridded Map):: https://cds.climate.copernicus.eu/#!/home
● Soil organic carbon (SOC) stock data: https://soilgrids.org/
If there are any questions about what exact variables should be used, use your
own judgment to make a choice.*

Comments on CDS functionality and data requests: https://cran.r-project.org/web/packages/ecmwfr/vignettes/cds_vignette.html


I will download this data using the package 'ecmwfr'
```{r}
lapply(c("ecmwfr","keyring"), library, character.only = TRUE)

ecmwfr::wf_set_key(user = "269199", key = "a107d4dc-ed19-46dc-ad85-390e61b50920", service = "cds")



```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
