---
title: "Lindsey's Climate Farmers Challenge"
output:
  html_document:
    df_print: paged
---

The work involves the use and analysis of gridded data (usually remote sensing or
model products), including land cover, climate and soil. You will need to download
and visualize relationships between these over a chosen region.

Load required packages
```{r}
# List of packages to load
packages_today <- c("tidyverse","ecmwfr","ggspatial","sf")

# Load the packages
lapply(packages_today, library, character.only = TRUE)

setwd("C:/Users/linds/Dropbox/Documents (Selective Sync Conflict)/Job Applications/Climate Farmers/CF-Challenge")

```



_**Step 1: Region Selection**_

*Decide on a specific region to work with. This region should be large enough to
include various land use types but small enough to keep computation times
reasonable. A country or region within a country for example.
The region can be defined using simple longitude and latitude limits or using a
geospatial file (e.g. shape file or geojson). Use this region to filter downloaded data
in the next steps.*


First, I load the required packages for this step (sf and rnaturalearth) to access shape files of country and regional boundaries. I will work with the state of California, USA.
```{r}
# Load the packages

lapply(c("sf","rnaturalearth"), library, character.only = TRUE)

# Get the natural earth data for U.S. states
california_sf <- ne_states(country = "united states of america", returnclass = "sf") %>% 
  filter(name == "California") %>% st_as_sf()
```


I confirm the correct geometry of the shape file and write to the computer, encoding with "UTF-8", as in the original file. 
```{r}
#Use base R to view state outline
plot(california_sf$geometry)

#Save shapefile
#st_write(california_sf, "california_sf.shp", driver = "ESRI Shapefile", encoding ="UTF-8")
```


I read back and plot the saved shape file to confirm correct writing to the drive.
```{r}
#Check file saved correctly by reading in shape file and plotting geometry
#test_shp<-st_read("california_sf.shp")
#plot(test_shp$geometry)
```


Finally, I plot on coordinates with ggplot2 to confirm the correct geographic position and begin building a graph for later visualisation.
```{r}
#Generate graph on coordinates to confirm position
ggplot(data = california_sf) +
  geom_sf() +
  ggtitle("California, USA") +
  theme(plot.title = element_text(hjust = 0.5)) +
  #xlab("") + ylab("") +
  geom_sf(color = "black", fill = "grey") 
```

_**Step 2: Data Acquisition**_

*Download data for your region for:
● Monthly evapotranspiration, temperature, and precipitation covering the
years 2000 to 2022 from (ERA5-Land):
https://cds.climate.copernicus.eu/#!/home
● Land cover classification maps for the year 2020 from Land Cover Classes
(Gridded Map):: https://cds.climate.copernicus.eu/#!/home
● Soil organic carbon (SOC) stock data: https://soilgrids.org/
If there are any questions about what exact variables should be used, use your
own judgment to make a choice.*

Comments on CDS functionality and data requests: https://cran.r-project.org/web/packages/ecmwfr/vignettes/cds_vignette.html


Ideally, I would have downloaded this data using the package 'ecmwfr'. First adding the keychain. Instead, for simplicity in completing the activity, I downloaded the data directly from the website using the built-in data navigator 
```{r}
lapply(c("ecmwfr","keyring","ggmap"), library, character.only = TRUE)

ecmwfr::wf_set_key(user = "269199", key = "a107d4dc-ed19-46dc-ad85-390e61b50920", service = "cds")


```

Data Acquisition

● Monthly evapotranspiration, temperature, and precipitation covering the
years 2000 to 2022 from (ERA5-Land):
https://cds.climate.copernicus.eu/#!/home


Examine file with the raster package:

I chose to download total evaporation, soil temperature layer 1, and total precipitation at 12:00. Midday measurements will capture the transpiration component of evapotranspiration. During the default 0:00 measurements, the stomata of C3 plants are closed; thus no transpiration takes places. 

Comments on use of midday measurements:
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8739081/

Shallow soils (layer 1) contain the organic horizon and are most dynamic in response to ambient temperature and precipitation.

I wrote a function to extract the data by climate variables, defining a brick for each one in the global environment. The new named brick files should be: stl1_brick, e_brick, tp_brick


```{r}
lapply(c("raster","ncdf4"), library, character.only = TRUE)

#read .nc file and convert to a brick - specify a variable
#varname: stl1, e, tp

nc_file_path <- "temp_evap_prec.nc"
variable_name <- "stl1"

read_var <- function(nc_file_path, variable_name) {

    var_name <- deparse(substitute(variable_name)) #unevaluated argument name, converted to character 
    
    clean_var_name <- gsub("\"", "", var_name) #remove double quotations
    
  # Read the NetCDF file for the specified variable
 #Use the cleaned variable name in the assign function
    assign(paste0(clean_var_name, "_brick"), brick(nc_file_path, varname = variable_name), envir = .GlobalEnv)

  #crop to CA and plot
  var_cropped <- crop(get(paste0(clean_var_name, "_brick")), extent(california_sf))
  var_masked <- mask(var_cropped, california_sf)
  plot(var_masked, main = paste0(clean_var_name, "_brick"))
}

read_var("temp_evap_prec.nc","stl1")
read_var("temp_evap_prec.nc","e")
read_var("temp_evap_prec.nc","tp")

```


● Land cover classification maps for the year 2020 from Land Cover Classes
(Gridded Map):: https://cds.climate.copernicus.eu/#!/home

Look at land
cover classes from 2020

```{r}
landcover <- raster("landcover.nc")
#crop to CA and plot
var_cropped <- crop(landcover, extent(california_sf))
cal_cover <- mask(var_cropped, california_sf)
plot(cal_cover)

summary(cal_cover)  

```
Interpret the land cover data. IPCC Classes on page 15:

https://datastore.copernicus-climate.eu/documents/satellite-land-cover/D5.3.1_PUGS_ICDR_LC_v2.1.x_PRODUCTS_v1.1.pdf

The most common classes are:

10 - Rainfed cropland
60 - Forest, Tree cover, broadleaved, deciduous, closed to open (> 15%)
110 - Grassland, Mosaic herbaceous cover (>50%) / tree and shrub (<50%)
120 - Shrubland

```{r}
library(sp)  #Load package for managing spatial data
unique(cal_cover$Land.cover.class.defined.in.LCCS)
hist(cal_cover$Land.cover.class.defined.in.LCCS)

```

For now, I am leaving out the soil carbon data due to time constraints. I may be able to return later to this task.


Step 3: Data Integration
Reproject the different rasters to match the one with lowest resolution, thus allowing joint analysis.

Land cover resolution: [1] 0.002777778 0.002777778 (degrees, longlat)

ERA5 climate data resolutions: [1] 0.1 0.1 (degrees, longlat)
```{r}
#Confirm resolution of each data set

res(cal_cover)
proj4string(cal_cover)

res(stl1_brick)
proj4string(stl1_brick)


```



Now we will reproject the landcover data to the lower resolution of the ERA5 climate data. The 'resample' function from the 'raster' package estimates a new value for the lower resolution data, based on a defined method. I chose 'mode', which is suitable for relatively preserving the categorial data of landcover classes.

```{r}
# Resample the higher resolution raster to the resolution of the lower resolution raster
cal_cover_res <- resample(cal_cover, stl1_brick, method = "ngb")
res(cal_cover_res)
plot(cal_cover_res)
unique(cal_cover_res)

```


Graph the high- and low-resolution histograms of land cover classes to check distribution in both projections. Note that a few rare land cover classes were lost in resampling, but the main categories are retained:

10 - Rainfed cropland
60 - Forest, Tree cover, broadleaved, deciduous, closed to open (> 15%)
110 - Grassland, Mosaic herbaceous cover (>50%) / tree and shrub (<50%)
120 - Shrubland

Unique values of land cover class are:
 [1]  10  11  30  40  70  80  90 110 120 130 160 180 190 200 210

```{r}

par(mfrow=c(1,2))
hist(cal_cover$Land.cover.class.defined.in.LCCS)
hist(cal_cover_res$Land.cover.class.defined.in.LCCS)
par(mfrow=c(1,1))
```


Step 4: Analysis and Visualization
Visualize the time series of climate and soil variables in the format of your choice,
aggregated separately for different land cover classes.

```{r}

head(cal_cover_res)
head(stl1_brick)


# Convert the brick to a data frame

stl1_df <- rasterToPoints(stl1_brick) %>% as.data.frame()
cover_df <- rasterToPoints(cal_cover_res) %>% as.data.frame()

full <- full_join(cover_df, stl1_df, by = c("x","y"))


# Plot the time series using ggplot2
ggplot(df_long, aes(x = Date, y = Soil_Temperature)) +
  geom_line() +
  ggtitle("Soil Temperature Time Series in California") +
  xlab("Date") +
  ylab("Soil Temperature") +
  theme_minimal()

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
